Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-07-20 paper.tex
.

In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions. The general method workflow is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters and relating these variations of values to the variations in the simulation results, we here introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. Initial spatial conditions are clustered to represent types of spaces ex-ante (for example: moonocentric or polycentric density grids), and the sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types. It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.

@@

Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}). In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability proportional to $P_i/\sum P_k$ to be added to patch $i$ with population $P_i$. At the end of each time step, this first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid bord effects such as a reflexion on the border of the world, border patches diffuse to the outside.  The model stops when a fixed total population is reached. Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf. figure \ref{fig:spatialGen}). Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
\comment{A detailler en incluant les equations de morphogenese}

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/spatialGen.png}
} \caption{Four examples of grids produced by the spatial generator. The lighter the red, the denser the area.} \label{fig:spatialGen} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

@@

In order to test for the influence of spatial initial conditions, we need a systematic method to compare phase diagrams. Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic. A solution is to use systematic quantitative procedures. Several potential methods could be used \comment[MLT]{Ici je me demande si l'on ne devrait pas simplement dire que l'on a pas trouvé de méthodes de comparaisons systématiques de diagrammes de phase dans la littérature SMA ou geographic modelling, et que du coup on propose de comparer la variance intra à la variance inter-grille selon la formule ... Et je garderais la précision sur les autres méthodes que l'on aurait pu choisir pour la section de discussion}: for example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters). The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools. Map comparison methods, popular in environmental sciences \comment[MLT]{Et pas en géo? Qu'est-ce que ces méthodes ont de particulier?}, provide numeral tools to compare two dimensional fields~\cite{visser2006map}. To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}. To keep it simple and as such methodological considerations are auxiliary to the main purpose of this paper, we propose an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability. More formally, the distance is given by

@@

where $\alpha \mapsto \left[\vec{x} \mapsto f_{\vec{\alpha}}\left(\vec{x}\right)\right]$ is the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a distance between probability distributions that can be taken for example as basic L2 distance \comment[MLT]{Euclidean? Je crois qu'il faut adapter le vocabulaire utilisé aux lecteurs d'EPB, qu'en pensez-vous?} or the Earth's Mover Distance \comment[MLT]{Si l'on mentionne les deux, ne va-t-on pas nous demander de tester l'effet du choix de l'une ou de l'autre sur nos résultats?}. For each values $\vec{\alpha_i}$, the phase diagram is seen as a random spatial field, facilitating the definition of variances and distance.
\comment[FL]{je pense que c'est un peu compliqué effectivement cette partie.}

@@

The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows. The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in~\cite{openshaw1983data}, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMole~\cite{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file. \comment[MLT]{Je me suis permis de bouger la partie sur le multi-modeling en fin de paragraphe, mais ai essayé d'en garder l'esprit tel quel!} In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modeling claimed by \cite{cottineau2016back}.

@@

We use our set of initial density grids in two spatial simulation models: Schelling and Sugarscape.

@@

\comment[FL]{il faudrait peut être davantage renforcer l'analyse a propos de figure 1 de si c'est plutôt les mu parameters ou les spatial parameters qui ont une grosse influence sur les résultats.}

@@

We obtain a very strong sensitivity to initial conditions, as the distribution of the relative distance to reference across grids ranges from 0.09 to 2.98 with a median of 1.52 and an average of 1.30. It means that in average, the model is more sensitive to meta-parameters than to parameters, and the relation variation can reach a factor of 3. We plot in Fig.~\ref{fig:sugarscape-distance} their distribution in a morphological space. The reduced morphological space is obtained by computing 4 raw indicators of urban form, namely Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and contextualization), and by reducing the dimension with a principal component analysis for which we keep the first two components (92\% of cumulated variance). The first measures a ``level of sprawl'' and of scattering, whereas the second measures aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation. It is confirmed by the behavior as a function of meta-parameters, as high values of $\alpha$ also yield high distance. In terms of model processes, it shows that congestion mechanisms induce rapidly higher levels of inequality.

@@

\comment{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}

@@

>>>>

In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions. The general method workflow is illustrated in Figure \ref{fig:method}. In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters and relating these variations of values to the variations in the simulation results, we here introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. Initial spatial conditions are clustered to represent types of spaces ex-ante (for example: monocentric or polycentric density grids), and the sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types. It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.

@@

Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}). In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$. More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel. At the end of each time step, this first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches). To avoid bord effects such as a reflexion on the border of the world, border patches diffuse to the outside.  The model stops when a fixed total population is reached. Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf. figure \ref{fig:spatialGen}). Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
\comment{A detailler en incluant les equations de morphogenese}[(JR) j'ai détaillé le processus, c'est suffisant ?]

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/spatialGen.png}
} \caption{Four examples of grids produced by the spatial generator. The lighter the red, the denser the area. Changing the growth rate $N$ allows to have more or less chaotic shapes corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.} \label{fig:spatialGen} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

@@

In order to test for the influence of spatial initial conditions, we need a systematic method to compare phase diagrams. Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic. A solution is to use systematic quantitative procedures. To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature. Several potential methods from other fields such as environmental science could be used, but we keep it simple and such methodological considerations are furthermore auxiliary to the main purpose of this paper. we propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability. More formally, the distance is given by

@@

where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$. $d$ is a functional distance that we take simply as Euclidian distance. The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.

\comment[JR]{j'ai simplifié au max}

@@

The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows. The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in~\cite{openshaw1983data}, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMole~\cite{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file. In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modeling claimed by \cite{cottineau2016back}.

@@

We use our set of initial density grids in two ``classical'' spatial simulation models: Schelling and Sugarscape.

@@

\comment[FL]{il faudrait peut être davantage renforcer l'analyse a propos de figure 1 de si c'est plutôt les mu parameters ou les spatial parameters qui ont une grosse influence sur les résultats.}[(JR) assez explicite pour sugarscape, j'ai rajouté un peu tout de même.]

@@

We obtain a very strong sensitivity to initial conditions, as the distribution of the relative distance to reference across grids ranges from 0.09 to 2.98 with a median of 1.52 and an average of 1.30. It means that in average, the model is more sensitive to meta-parameters than to parameters, and the relation variation can reach a factor of 3. We plot in Fig.~\ref{fig:sugarscape-distance} their distribution in a morphological space. The reduced morphological space is obtained by computing 4 raw indicators of urban form, namely Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and contextualization), and by reducing the dimension with a principal component analysis for which we keep the first two components (92\% of cumulated variance). The first measures a ``level of sprawl'' and of scattering, whereas the second measures aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation. It is confirmed by the behavior as a function of meta-parameters, as high values of $\alpha$ also yield high distance. In terms of model processes, it shows that congestion mechanisms induce rapidly higher levels of inequality. To put these results in perspective of our workflow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters in average greater than the sensitivity to model parameters.

@@

\comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]

@@

\paragraph{Comparing phase diagrams}

Comparing phase diagrams is as we saw not straightforward, and further developments of our method imply testing alternative methods for this particular point. For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters). The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools. Map comparison methods, popular in environmental sciences \comment[MLT]{Et pas en géo? Qu'est-ce que ces méthodes ont de particulier?}, provide numeral tools to compare two dimensional fields~\cite{visser2006map}. To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}. Note that with the intuitive index we used, more subtle distances such as the Earth's Mover Distance can be used.
